# Introduction {#intro}


## Opportunities in GIS

Significant opportunities exist in the development of GIS. 

## Polygons in GIS are not manifolds

What is a polygon? 

Wikipedia: 

* a polygon is a plane figure that is bounded by a finite chain of straight line segments closing in a loop to form a closed chain or circuit

```{r}
p <- cbind(x = c(0, 0, 0.75, 1, 0.5, 0.8, 0.69, 0), 
           y = c(0, 1, 1, 0.8, 0.7, 0.6, 0))
plot(p)
lines(p)
```

Couldn't be simpler. Note that we visited 0,0 twice, when we first started and at the final destination. What if we left out the last 0,0?

```{r}
npoints <- nrow(p)
plot(p)
lines(p[seq(npoints - 1), ])
```

Well it's obvious, we don't have a polygon so we just end up with a line. But what happened in the code `r plot(p)` - we don't see two points, though that is just because they both were plotted but are indistinguishable. Let's return to this point later. 


What if we distinguish inside and outside of this part of the plane? 

```{r}
plot(p)
polygon(p)
```
What happened? We have a close loop of line segments. 

```{r}
p1 <- p[-nrow(p), ]  ##remove the final 0,0 point
plot(p1)
polygon(p1)
```

Hmm no difference. 

I'm not convinced that this is anything but a set of lines, so let's set the background colour first. 

```{r}
op <- par(bg = "lightblue")
plot(p1)
polygon(p1)
```

Still not convinced. We have to tell `polygon()` to fill the space it contains. 

```{r}
plot(p1)
polygon(p1, col = "grey")
par(op)
```


What is a line? 

A line is a finite chain of straight line segments. 



