---
title: "Polygons in R"
author: "Michael Sumner"
date: "2016-03-04"
output:
  ioslides_presentation:
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, size = 0.5)
```

## What is a polygon? 

> **Wikipedia: 
> a polygon is a plane figure that is bounded by a finite chain of straight 
> line segments closing in a loop to form a closed chain or circuit**

Straight line segments? 

Boundary coordinates connected by a traced "line"? 

Does it matter? 

## The simplest polygon

```{r,echo=FALSE}
tri <- cbind(c(0, 1, 1), c(0, 0, 1))
op <- par(mfrow = rep(2, 2), xpd = NA)
plot(tri, type = "n", axes = FALSE, xlab = "", ylab = "", main = "plane figure")
polygon(tri, col = "grey", border = NA)
plot(tri, type = "n", axes = FALSE, xlab = "", ylab = "", lty = 2, main = "straight line segments")

polygon(tri, col = "grey", border = NA, density = 10, angle = 65)
polygon(tri, col = "grey", border = NA, density = 10, angle = 10)

#arrows(tri[c(1, 2, 3), 1], tri[c(1, 2, 3), 2], tri[c(2, 3, 1), 1], tri[c(2, 3, 1), 2], length = 0.25, lty = 1, lwd = 2, angle = 15)
arrows(tri[c(1, 2, 3), 1], tri[c(1, 2, 3), 2], tri[c(2, 3, 1), 1], tri[c(2, 3, 1), 2], length = 0, lty = 1, lwd = 2, col = c("#1B9E77", "#D95F02", "#7570B3"))

plot(tri, type = "n", axes = FALSE, xlab = "", ylab = "", lty = 2, main = "boundary coordinates")
polygon(tri, col = "grey", border = NA)
points(tri, pch = 1, cex = 1.2)

plot(tri[c(1:3, 1), ], type = "l", axes = FALSE, xlab = "", ylab = "", lty = 2, main = "boundary line")
par(op)
```

Aren't these all the same thing? 


## Connected vertices

```{r,echo=FALSE}
xl <- c(-0.2, 1.35); yl <- c(-0.2, 1.2)
p <- cbind(x = c(0, 0, 0.75, 1,   0.5, 0.8, 0.69, 0), 
           y = c(0, 1, 1,    0.8, 0.7, 0.6, 0,    0))
rownames(p) <- sprintf("pt%i", seq(nrow(p)))
#print(p)

#op <- par()
layout(matrix(c(1, 1, 1, 2, 2, 2, 2, 2, 2), ncol = 3))
plot(0, type = "n", axes = FALSE, ylim = c(0, 8), xlim = c(0, 1), 
     xlab = "",  ylab = "")
text(rep(0.2, nrow(p)), nrow(p):1, labels = sprintf("%s:  (%1.2f,%1.2f)", rownames(p), p[,1], p[,2]), cex = 0.9, pos = 4)


plot(p, xlim = xl, ylim = yl)
lines(p)
text(p, labels = sprintf("%s:  (%1.2f,%1.2f)", rownames(p), p[,1], p[,2]), cex = 0.75,
     pos = c(1, 3, 3, 4, 3, 4, 1, 3), col = c(rep("black", nrow(p)-1), "firebrick"))
#par(op)
```

Note that we visited 0,0 twice, when we first started as "pt1" and at the final destination as "pt8". What if we left out the last 0,0?



```{r}
npoints <- nrow(p)
plot(p, xlim = xl, ylim = yl)
lines(p[seq(npoints - 1), ])
```

Well it's obvious, we don't have a polygon so we just end up with a line. But what happened in the code `r plot(p)` - we don't see two points, though that is just because they both were plotted but are indistinguishable. Let's return to this point later. 


What if we distinguish inside and outside of this part of the plane? 

```{r}
plot(p, xlim = xl, ylim = yl)
polygon(p)
```

What happened? We have a closed loop of line segments. 

```{r}
p1 <- p[-nrow(p), ]  ##remove the final 0,0 point
plot(p1, xlim = xl, ylim = yl)
polygon(p1)
```

Hmm no difference. 

I'm not convinced that this is anything but a set of lines, so let's set the background colour first, in case it's a white polygon on a white background. 

```{r}
op <- par(bg = "lightblue")
plot(p1, xlim = xl, ylim = yl)
polygon(p1)
```

We are seeing right through this polygon. We have to tell `polygon()` to fill the space it contains with a colour (which might be white). 

```{r}
plot(p1, xlim = xl, ylim = yl)
polygon(p1, col = "grey")
par(op)
```

What if we have two of these kinds of objects? What goes wrong here? 

```{r}
p1 <- cbind(x = c(0, 0, 0.75, 1,   0.5, 0.8, 0.69, 0), 
           y = c(0, 1, 1,    0.8, 0.7, 0.6, 0,    0))
p2 <- cbind(x = c(0.2, 0.2, 0.3, 0.5, 0.5, 0.2), 
            y = c(0.2, 0.4, 0.6, 0.4, 0.2, 0.2))

pp <- rbind(p1, p2)

plot(pp, xlim = xl, ylim = yl, main = "yikes!")

polygon(rbind(p1, p2), col = "grey")

```

We have defined a *path* and we've told the polygon-drawing-tool to follow that path and fill space with a particular rule. (We what?). 

One way to fix this is to plot the two parts separately, but we've cheated since the white "hole" is not a hole - we can no longer see the points on the inner ring. 

```{r}

plot(pp, xlim = xl, ylim = yl)

polygon(p1, col = "grey")
polygon(p2, col = "white")
```

The proper way to do this is is to separate the parts of the path where we don't want the path to stay connected. But, here `polygon` doesn't understand that it was meant to draw it as a hole. 

```{r}
pph <- rbind(p1, NA, p2)

plot(pph, xlim = xl, ylim = yl)

polygon(pph, col = "grey")


```

The only holes that `polygon` knows about are the ones determined by the evenodd rule. 

```{r}
pph1 <- rbind(p1, p2 + 0.35)

plot(pph1, xlim = xl, ylim = yl)

polygon(pph1, col = rgb(0, 0, 0, 0.2), fillOddEven = FALSE)


```


Clockwise or counter-clockwise for coordinates?  It depends, there is the winding rule (it matters) and the evenodd rule (it doesn't matter), but those rules only apply for a polygon-path method if the polygon is self-intersecting. In GIS, it depends on the format, assumptions within a program, etc. 

Individual polygons. 

Polygons that are holes? Or just overlapping? 

Winding or the even-odd rule? 

Formats for polygons: 
 coordinates separated by a special value (i.e. NA)
 sp Spatial classes
 spatstat, maps, WKT, WKB, shapefiles, MapInfo files (TAB or MIF)
 triangulations, constrained Delaunay, ear-clipping
 
## Limits in R

polygon, and polypath can only plot one colour at a time
- in practice this means sp draws every object separately (still pretty fast though)

polypath can draw holes where we separate a hole from its outer ring by missing values, and use a rule for how to draw
How far can we take this in R? 

What about polygons on the sphere? Polygons on a topographic surface? 

What is interesting about this polygon? It has a slight complex shape, i.e. it is not identical to the convex hull of the points. It also has 7 unique vertices, so is marginally complicated. Whas is interesting about these polygons? 

```{r}
library(maptools)
data(wrld_simpl)
```




```{r}
pph <- rbind(p1, NA, p2)

plot(pph, xlim = xl, ylim = yl)

polypath(pph, col = "grey", rule = "evenodd")


```

## polypath with no separating
```{r}

plot(pp, xlim = xl, ylim = yl, main = "yikes")
polygon(rbind(p1, p2), col = "grey")

op <- par(mfrow = c(2, 2))
plot(pp, xlim = xl, ylim = yl, main = "winding, clock")
polypath(rbind(p1, p2), col = "grey", rule = "winding")

plot(pp, xlim = xl, ylim = yl, main = "winding, counter")
polypath(rbind(p1[nrow(p1):1, ], p2[nrow(p2):1, ]), col = "grey", rule = "winding")

plot(pp, xlim = xl, ylim = yl, main = "winding, clock/counter")
polypath(rbind(p1, p2[nrow(p2):1, ]), col = "grey", rule = "winding")

plot(pp, xlim = xl, ylim = yl, main = "winding, counter/clock")
polypath(rbind(p1[nrow(p1):1, ], p2), col = "grey", rule = "winding")

par(op)

op <- par(mfrow = c(2, 2))
plot(pp, xlim = xl, ylim = yl, main = "evenodd, clock")
polypath(rbind(p1, p2), col = "grey", rule = "evenodd")

plot(pp, xlim = xl, ylim = yl, main = "evenodd, counter")
polypath(rbind(p1[nrow(p1):1, ], p2[nrow(p2):1, ]), col = "grey", rule = "evenodd")

plot(pp, xlim = xl, ylim = yl, main = "evenodd, clock/counter")
polypath(rbind(p1, p2[nrow(p2):1, ]), col = "grey", rule = "evenodd")

plot(pp, xlim = xl, ylim = yl, main = "evenodd, counter/clock")
polypath(rbind(p1[nrow(p1):1, ], p2), col = "grey", rule = "evenodd")
par(op)
```

## polypath with NA separating
```{r}

op <- par(mfrow = c(2, 2))
plot(pp, xlim = xl, ylim = yl, main = "winding, clock")
polypath(rbind(p1, NA, p2), col = "grey", rule = "winding")

plot(pp, xlim = xl, ylim = yl, main = "winding, counter")
polypath(rbind(p1[nrow(p1):1, ], NA, p2[nrow(p2):1, ]), col = "grey", rule = "winding")

plot(pp, xlim = xl, ylim = yl, main = "winding, clock/counter")
polypath(rbind(p1, NA, p2[nrow(p2):1, ]), col = "grey", rule = "winding")

plot(pp, xlim = xl, ylim = yl, main = "winding, counter/clock")
polypath(rbind(p1[nrow(p1):1, ], NA, p2), col = "grey", rule = "winding")

par(op)

op <- par(mfrow = c(2, 2))
plot(pp, xlim = xl, ylim = yl, main = "evenodd, clock")
polypath(rbind(p1,NA,  p2), col = "grey", rule = "evenodd")

plot(pp, xlim = xl, ylim = yl, main = "evenodd, counter")
polypath(rbind(p1[nrow(p1):1, ],NA,  p2[nrow(p2):1, ]), col = "grey", rule = "evenodd")

plot(pp, xlim = xl, ylim = yl, main = "evenodd, clock/counter")
polypath(rbind(p1, NA, p2[nrow(p2):1, ]), col = "grey", rule = "evenodd")

plot(pp, xlim = xl, ylim = yl, main = "evenodd, counter/clock")
polypath(rbind(p1[nrow(p1):1, ],NA,  p2), col = "grey", rule = "evenodd")
par(op)
```
